Global patterns of forest autotrophic carbon fluxes

### 2020-01-17 manuscript/manuscript draft.Rmd
finished main doc figures review
working on Helene's comments

###Abstract
[**very rough start:**]
..... Here, we drew upon # records from the Global Forest Carbon Database (ForC), representing all major forest types and the nine most significant forest autotrophic carbon flux (FACF) variables, to comprehensively explore how C cycling in mature, undisturbed forests varies with latitude and climate on a global scale. We show that, across all FACF variables analyzed, C cycling decreases linearly with latitude --a finding that confirms multiple previous studies and contradicts the idea that productivity of temperate forests rivals that of tropical forests. The FACF variables increase in proportion to one another, with no differences in allocation detectable at this global scale, but differed in that latitude explained a lower proportion of variation among subsidiary fluxes (in particular, woody aboveground productivity and belowground net primary productivity, $BNPP$). Climate explained a significant proportion (#-#%) of variation in all C fluxes analyzed (less for subsidiary fluxes), with temperature variables in general and mean annual temperature ($MAT$) in particular being the best predictors of FACF on this global scale. While other climate variables (*e.g.*, **XX**) displayed significant correlation with FACF, none of them improved significantly over the explanatory power $MAT$. The effects of temperature were modified by moisture availability, with reduced FACF under hot and dry conditions and sometimes under very high precipitaiton (especially for $BNPP$). FACF declined with temperature seasonality, but growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of carbon fluxes. These findings clarify the big picture of how FACF varies with latitude and climate on a global scale. As we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

@@

**Forests play a critical role in regulating atmospheric CO~2~ and climate** (**Bonan**). The total amount of CO~2~ cycling through Earth's forests each year (*i.e.*, total gross primary production, $GPP$) is more than five times anthropogenic fossil fuel emissions, and net sequestration (~2.4 GtCyr^-1^) offsets roughly 30% of these emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, with consequences for global climate, there is increasing recognition that  protection and management of forest resources will play an important role in mitigating climate change (** REFS- e.g., IPCC 1.5C and land reports**). Carbon fixation, allocation, and metabolism by trees set the basis for energy and material flows in forest ecosystems and define their interactions with Earth's changing climate.

**Forests' autotrophic carbon flux (FACF, bbl:define) influences all organic matter stocks in ecosystems and is linked to cycling of energy, water, and nutrients** (**REFS; *Piao et al. 2010. Forest annual carbon cost: a global‐scale analysis of autotrophic respiration. Ecology 91(3), 652-661* **). Forest productivity sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity ([Waide et al. 2009](https://www.annualreviews.org/doi/abs/10.1146/annurev.ecolsys.30.1.257)**REFS**)--but probably not individual metabolic rate (**Anderson & Jetz 2005, but check more recent literature**). However, our ability to draw general macroscopic conclusions regarding global variation in multiple productivity variables with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables (*e.g.*, combining net primary productivity records with and without belowground components); and typically consider only one or a few variables at a time.

**The productivity of forests varies with latitude**, showing a general trend of decreasing productivity with latitude [@beer_terrestrial_2010; @jung_global_2011]. Studies agree that productivity is lowest in the boreal regions, and increases into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity (GPP) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011];[Bagdley et al. 2019](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.14729)). This is corroborated by analysis of site-level GPP measurements, which appear to reach their highest levels in tropical forests [@luyssaert_co_2007]. In contrast, there is evidence that the highest values of net primary productivity (NPP) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], although other studies find NPP is highest in the tropics, showing a decrease with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity (ANPP), finding evidence of a weak negative relationship between ANPP and latitude [@huston_global_2009; @gillman_latitude_2015].

**Climate is a significant driver of productivity across broad spatial scales** [@cleveland_relationships_2011]. The majority of studies have focused on exploring the relationships between productivity and mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. These variables have the advantage that they describe broad trends in temperature and water availability, and therefore capture a lot of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with productivity [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of productivity across studies. Various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while GPP increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in GPP saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that NPP increases linearly with temperature. bbl:maybe remove following sentences The influence of these climate variables on productivity is further complicated by the possibility of interactive effects occurring between them. Taylor et al. [-@taylor_temperature_2017] showed that increased MAP had a negative effect at low MAT, but a positive effect at high levels of MAT, and vice versa, suggesting that both temperature and water availability are important in explaining productivity.

**Primary productivity can be influenced by many other factors as well**, which often act across a range of scales, and may show interactive effects with each other [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity (Liang et al. 2019. Positive biodiversity-productivity relationship predominant in global forests. Science 354, Issue 6309, aaf8957), management [@simova_enigma_2017]; nutrient availability [@aragao_above-_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact forest productivity. On a global scale, we expect that productivity is most strongly influenced by broad climatic gradients. There is evidence that productivity also responds to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005] in potentially significant ways. Furthermore, MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and productivity is actually a factor of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between productivity and MAT, indicating that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on productivity.

@@

To investigate the potential interactive effects of climate variables on carbon fluxes, multivariate models were also specified. To ensure that models were biologically meaningful, the terms included in the models tested built on results from the univariate models. Modelling of individual climate variables identified that the best predictors of carbon fluxes were variables related to temperature. We therefore decided to include mean annual temperature as a term in all multivariate models. We first modelled the interaction effect between mean annual temperature and mean annual precipitation, in order to capture climate variation along the axes of temperature and water availability. Models were tested for a significant interactive effect and a significant additive effect. We then explored whether any other climate variable, in combination with mean annual temperature, could significantly improve on the combination of mean annual temperature and mean annual precipitation. In specifying the range of models to test, climate variables which were strongly correlated with temperature were dropped, in order to capture the greatest range of variation in climate. For each possible pairing of climate variables, two models were specified: a model with the two climate variables showing an additive effect; and a model with the two climate variables showing an interactive effect. As before, plot nested within geographic area was included as a random effect. Altitude was not considered. AIC values were calculated for the models, and used to compare models. Models were considered to be significantly better than the baseline MAT*MAP model if:  
  i) the AIC value of the model was smaller than the AIC value of the baseline model by >2  
  ii) the r-squared value of the model was larger than the r-squared value of the baseline model by >5

@@

All major carbon fluxes increased linearly with decreasing latitude (fig. 2).

>>>>

###Abstract
[**very rough start:**]
..... Here, we drew upon # records from the Global Forest Carbon Database (ForC), representing all major forest types and the nine most significant forest autotrophic carbon flux (FACF) variables, to comprehensively explore how C cycling in mature, undisturbed forests varies with latitude and climate on a global scale. We show that, across all FACF variables analyzed, C cycling decreases linearly with latitude --a finding that confirms multiple previous studies and contradicts the idea that productivity of temperate forests rivals that of tropical forests. The FACF variables increase in proportion to one another, with no differences in allocation detectable at this global scale, but differed in that latitude explained a lower proportion of variation among subsidiary fluxes (in particular, woody aboveground productivity and belowground net primary productivity, $BNPP$). Climate explained a significant proportion (#-#%) of variation in all C fluxes analyzed (less for subsidiary fluxes), with temperature variables in general and mean annual temperature ($MAT$) in particular being the best predictors of FACF on this global scale. While other climate variables (*e.g.*, **XX**) displayed significant correlation with FACF, none of them had significantly better explanatory power than $MAT$. The effects of temperature were modified by moisture availability, with reduced FACF under hot and dry conditions and sometimes under very high precipitaiton (especially for $BNPP$). FACF declined with temperature seasonality, but growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of carbon fluxes. These findings clarify the big picture of how FACF varies with latitude and climate on a global scale. As we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

@@

(**revisit these first two paragraphs.**)

**Forests play a critical role in regulating atmospheric CO~2~ and climate** (**Bonan**). Approximately **##** Gt of CO~2~ cycle through Earth's forests each year (*i.e.*, total gross primary production, $GPP$; **REF-KAT**), and their net C sequestration (~2.4 GtCyr^-1^) offsets roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving the global climate crisis, there is increasing recognition that  protection and management of forest resources will play an important role in mitigating climate change (** REFS- e.g., IPCC 1.5C and land reports**). Carbon fixation, allocation, and metabolism by trees set the basis for energy and material flows in forest ecosystems and define their interactions with Earth's changing climate.

**Forests' autotrophic carbon flux (FACF, bbl:define) influences all organic matter stocks in ecosystems and is linked to cycling of energy, water, and nutrients** (**REFS; *Piao et al. 2010. Forest annual carbon cost: a global‐scale analysis of autotrophic respiration. Ecology 91(3), 652-661* **). Forest productivity sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity ([Waide et al. 2009](https://www.annualreviews.org/doi/abs/10.1146/annurev.ecolsys.30.1.257)**REFS**). However, our ability to draw general macroscopic conclusions regarding global variation in multiple productivity variables with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables (*e.g.*, combining net primary productivity records with and without belowground components); and typically consider only one or a few variables at a time.

**The productivity of forests varies with latitude**, showing a general trend of decreasing productivity with latitude [@beer_terrestrial_2010; @jung_global_2011]. Studies agree that productivity is lowest in the boreal regions, and increases into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity (GPP) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011];[Bagdley et al. 2019](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.14729)). This is corroborated by analysis of site-level GPP measurements, which appear to reach their highest levels in tropical forests [@luyssaert_co_2007]. {**condense the previous two sentences into one-- both primarily eddy flux**} In contrast, some studies suggest that the highest values of net primary productivity (NPP) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find NPP highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity (ANPP), finding evidence of a weak negative relationship between ANPP and latitude [@huston_global_2009; @gillman_latitude_2015].

**Climate is a significant driver of productivity across broad spatial scales** [@cleveland_relationships_2011]. The majority of studies have focused on exploring the relationships of productivity to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. These variables have the advantage that they describe broad trends in temperature and water availability, and therefore capture a lot of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with productivity [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of productivity across studies. Various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while GPP increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in GPP saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that NPP increases linearly with temperature. bbl:maybe remove following sentences The influence of these climate variables on productivity is further complicated by the possibility of interactive effects occurring between them. Taylor et al. [-@taylor_temperature_2017] showed that increased MAP had a negative effect at low MAT, but a positive effect at high levels of MAT, and *vice versa* {**Helene found this confusing**}, suggesting that both temperature and water availability are important in explaining productivity.

**Primary productivity can be influenced by many other factors as well**, which often act across a range of scales, and may show interactive effects with each other [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity (Liang et al. 2019. Positive biodiversity-productivity relationship predominant in global forests. Science 354, Issue 6309, aaf8957), management [@simova_enigma_2017]; nutrient availability [@aragao_above-_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact forest productivity. On a global scale, we expect that productivity is most strongly influenced by broad climatic gradients. There is evidence that productivity also responds to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005] in potentially significant ways. Furthermore, MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and productivity is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between productivity and MAT, suggesting that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on productivity.

@@

To investigate the potential interactive effects of climate variables on carbon fluxes, multivariate models were also specified. To ensure that models were biologically meaningful, the terms included in the models tested built on results from the univariate models. Modelling of individual climate variables identified that the best predictors of carbon fluxes were variables related to temperature. We therefore decided to include mean annual temperature as a term in all multivariate models. We first modelled the interaction effect between mean annual temperature and mean annual precipitation, in order to capture climate variation along the axes of temperature and water availability. Models were tested for a significant interactive effect and a significant additive effect. We then explored whether any other climate variable, in combination with mean annual temperature, could significantly improve on the combination of mean annual temperature and mean annual precipitation. In specifying the range of models to test, climate variables which were strongly correlated with temperature were dropped, in order to capture the greatest range of variation in climate. For each possible pairing of climate variables, two models were specified: a model with the two climate variables showing an additive effect; and a model with the two climate variables showing an interactive effect. As before, plot nested within geographic area was included as a random effect. Altitude was not considered. AIC values were calculated for the models, and used to compare models. Models were considered to be significantly better than the baseline MAT*MAP model if:  
  i) the AIC value of the model was smaller than the AIC value of the baseline model by >2  
  ii) the r-squared value of the model was larger than the r-squared value of the baseline model by **>5 (???- not possible)**

@@

All major carbon fluxes decreased linearly with latitude (Fig. 2).
